---
title: "mergecalib: within-province cell merging and proportion-interval calibration"
author: "Kunxiang Ma"
date: "2026-06-22"
output:
  html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{mergecalib: cell merging and interval calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# 1. The problem addressed

`mergecalib` operates on data whose rows are demographic cells. Each cell is
defined by province, sex, urban/rural status, age group, and education level,
and carries a sample size, a total survey weight, and the counts of four outcome
grades A, B, C, and D. The package merges original cells **only within a
province** and adjusts the weighted proportions of provincial and national
populations through the common unit weight of each merged cell.

# 2. The merged weight

If a final cluster $C$ contains several original cells, then:

$$N_C=\sum_{i\in C}n_i,\qquad W_C=\sum_{i\in C}W_i$$

All observations in the cluster share the common unit weight:

$$\bar w_C=\frac{W_C}{N_C}$$

A candidate cluster must therefore satisfy $N_C>0$.

# 3. The hard constraint on zero-sample cells

The package enforces the following rules:

1. positive-sample cells may stand alone;
2. zero-sample cells may **not** stand alone;
3. every zero-sample cell must join a positive-sample candidate cluster in the
   same province;
4. each province is given at least one deterministic zero-sample absorption
   partition;
5. after solving, the condition `final_n > 0` is re-checked for every cluster;
6. if an entire province has total sample size 0, a structural infeasibility is
   reported immediately.

# 4. Target intervals

For a target population $h$ and grade $k$, we require:

$$L_{hk}\leq P_{hk}\leq U_{hk}$$

Because a candidate cluster's unit weight and its contribution to each target
become constants once the cluster is generated, the proportion bounds turn into
linear constraints and are handed to the mixed-integer linear program.

# 5. Quick example

```{r eval=FALSE}
library(mergecalib)
mergecalib_agree(TRUE)  # acknowledge the disclaimer for this session

spec <- merge_spec(
  level_orders = list(
    age = c("18-39", "40+"),
    education = c("low", "high")
  )
)

dat <- example_merge_data()
targets <- example_merge_targets(dat, spec)

fit <- fit_merge_calibration(dat, targets, spec)
print(fit)
summary(fit)
```

# 6. Wildcard rules in the target table

The province, sex, urban, age, and education fields of the target table may hold
a concrete value or `NA` / `"*"`. For example:

- provincial target: fill in the province, leave the other dimensions `NA`;
- national-male target: province `NA`, sex set to "M";
- national cross-classified target: province `NA`, with sex, urban, age, and
  education all set.

# 7. Candidate-expansion levels

By default the search proceeds through four levels: local pairwise merges, local
small clusters, broader candidates, and the widest candidates. The space is only
widened when the narrower candidates are infeasible, balancing speed and
feasibility.

# 8. Lexicographic optimisation

Once the targets are satisfied strictly, the package minimises, in order:

1. the reweighted (moved) sample size;
2. the demographic distance of merged cells;
3. the heterogeneity of the ABCD distribution;
4. the change in unit weight;
5. the number of merges.

The optimal value of each stage becomes a constraint for the next, so a
higher-priority objective is never sacrificed because of a subjective weighting.

# 9. Infeasible targets and the minimum relaxation

If the original targets are infeasible, the package re-solves at fixed
relaxation amounts and uses a binary search to find the minimum uniform
relaxation. The result is stored in:

```{r eval=FALSE}
fit$delta
```

# 10. Outputs and audit

```{r eval=FALSE}
cell_map(fit)
merge_plan(fit)
final_cells(fit)
target_results(fit)
province_results(fit)
national_cell_results(fit)
audit_merge_fit(fit)
```

`audit_merge_fit()` checks: final zero-sample clusters, the exact one-time
cover, sample-size conservation, weight conservation, the effective target
intervals, and the MILP constraint residuals.

# 11. Performance notes

For the common scale of around 100 original cells per province, start with the
default candidate levels. Only widen `max_neighbors`, `max_cluster_size`, and
`max_combinations_per_cell` when the targets are especially tight or the cells
are very heterogeneous. Solving is single-threaded by default to guarantee
determinism.

# 12. Cases that merging cannot solve

The following situations raise an explicit error:

- a province with total sample size 0;
- a target population with zero total observed sample;
- a row whose ABCD counts do not sum to the sample size;
- a positive-sample cell whose total weight is not positive;
- a self-contradictory target interval;
- no solution even within the allowed candidate space and maximum relaxation.
